Computer Assisted Verbal Autopsy: Comparing Large Language Models to Physicians for Assigning Causes to 6939 Deaths in Sierra Leone from 2019-2022

### 2025-06-18 manuscript/manuscript.tex
Working on prabhat's feedback 2
Cleaner plots

\title[Towards Computer Assisted Verbal Autopsy: A Case Study Comparing Large Language Models to Physicians for Assigning Causes to 6939 Deaths in Sierra Leone from 2019-2022]{Towards Computer Assisted Verbal Autopsy: A Case Study Comparing Large Language Models to Physicians for Assigning Causes to 6939 Deaths in Sierra Leone from 2019-2022}

@@

\textbf{Methods:} 6939 VA records in Sierra Leone from 2019 to 2022 were used to compare four computer models, GPT-3.5, GPT-4, InterVA-5, and InSilicoVA, to physician COD assignment at population and individual levels. Narratives were used for GPT-3.5/4, while questionnaires were used for InterVA-5/InSilicoVA. COD assignments were grouped into general COD categories consisting of 19, 10, and 7 categories for adult, child, and neonatal age groups. Cause Specific Mortality Fraction (CSMF) accuracy and Partial Corrected Concordance (PCCC) were used to compare models to physicians at population and individual levels respectively. CSMF and PCCC were evaluated overall and by COD, age group, and age ranges.

@@

\textbf{Conclusion:} All models performed well at the population level, while GPT-3.5/4 and InSilicoVA performed well at the individual level for some CODs. GPT models have yet to replace physician coding, but made improvements over InSilicoVA and InterVA-5. Our research lays the foundation for future work in computer assisted VA, where physicians utilize alternative COD assignments from computer models to help reduce ill-defined codes and physician disagreement.
}

@@

In 2019, 41 million people die prematurely from noncommunicable diseases every year, accounting for 74\% of all deaths globally \citep{worldhealthorganizationNonCommunicableDiseases2019}. Most of these deaths are preventable, but require adequate resource allocation, guided by evidence, to implement effective interventions and policies that target populations at risk \citep{benzigerGlobalBurdenDisease2016}. Thus, reliable counts and diagnoses of deaths enable decision makers to identify populations at risk to save lives and reduce premature deaths worldwide \citep{lawnMillionNeonatalDeaths2010,lassiCommunityBasedIntervention2015,liuExcessMortalityPersons2017,ewigCommunityacquiredPneumoniaEmergency2011}. However, most low-income countries do not have data on deaths or have registered less than half of the deaths in their country, with an even fewer 8\% of these registered deaths having a Cause of Death (COD) recorded \citep{worldhealthorganizationSCOREHealthData2021}. To fill this gap in death registrations, an alternative method known as Verbal Autopsy (VA) is used to collect data on deaths and determine their likely causes at scale \citep{desavignyIntegratingCommunitybasedVerbal2017,thomasVerbalAutopsyHealth2018,rampatigeSystematicReviewStatistics2014}, outside of traditional healthcare facilities where over half of deaths occur at home \citep{adairWhoDiesWhere2021}.

@@

Recently, Large Language Models (LLM), leveraging massive datasets and deep learning approaches, have made advances in performing a variety of Natural Language Processing (NLP) tasks using free-text, such as question answering, code generation, and even medical diagnosis \citep{changSurveyEvaluationLarge2023,lundChattingChatGPTHow2023,svyatkovskiyIntelliCodeComposeCode2020,hauptAIGeneratedMedicalAdvice2023}. In 2022, a widely-available LLM called ChatGPT was released by OpenAI with capabilities of answering natural language text inquiries using training data up to September 2021. ChatGPT-3 was based on several Generative Pre-trained Transformer (GPT) models between 2018 to 2020, namely GPT-1 to GPT-3, which had notable differences in training data sizes of 5 gigabytes to 45 terabytes from web sources that resulted in 117 million to 175 billion parameter models \citep{wuBriefOverviewChatGPT2023}. In March 2023, ChatGPT-4 was released with human-level performance on various professional and academic exams and benchmarks that outperformed ChatGPT-3 \citep{openaiGPT4TechnicalReport2023}. Given the limited usage of free-text open narratives in computer models for determining CODs, and recent advances in LLMs that leverage natural language text prompts, we conducted a case study with Sierra Leone deaths from VA in 2019 to 2022 to compare four models, GPT-3.5, GPT-4, InterVA-5, and InSilicoVA, to physicians for determining CODs.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{fig-methods-brief.pdf}
\caption{Case study methods.}\label{fig-methods-brief}
\end{figure}

@@

The performance of the four models were evaluated with metrics at the population and individual level by comparing their CGHR-10 COD outputs for 6939 records. Cause Specific Mortality Fraction (CSMF) accuracy was used to evaluate models on the population level (see Appendix \ref{app-methods-csmfa}), while Partial Chance Corrected Concordance (PCCC) was used to evaluate models on the individual level (see Appendix \ref{app-methods-pccc}) \citep{murrayRobustMetricsAssessing2011}. Both CSMF accuracy and PCCC metrics are between 0 and 1 with 0 indicating low performance and 1 indicating perfect performance at the population and individual level respectively. As model performance can vary across ages and specific causes \citep{benaraEvaluationMethodsAssigning2023,murrayUsingVerbalAutopsy2014,setelValidityVerbalAutopsy2006}, the CSMF accuracy and PCCC metrics were compared for each model overall, by age group (adult, child, neonatal), by CGHR-10 COD codes, and across age ranges. For each of the adult and child age groups, metrics were calculated for five-year age ranges for records with ages at death of one-year or older and five-month age ranges for 28 days or older. For the neonatal age group, the age ranges of 0-6 days and 7-27 days were used. See Appendix \ref{app-methods-eval} for more details on performance metrics and evaluation strategy for comparing each model.

@@

This section details the performance results of GPT-3.5, GPT-4, InterVA-5, and InSilicoVA models for assigning CGHR-10 CODs after applying the methods in Section \ref{methods}. GPT-4 performed the best overall at 0.61 PCCC followed by GPT-3.5 at 0.56 PCCC. GPT-4 also had the highest PCCC for most age ranges and CODs across the adult (12 to 69 years), child (28 days to 11 years), and neonatal (under 28 days) age groups with GPT-3.5, InterVA-5, and InSilicoVA having higher PCCC values for a few age ranges and CODs. Overall performance results are seen in Section \ref{results-overall}, and performance by adult, child, and neonatal records are seen in Sections \ref{results-adult}, \ref{results-child}, and \ref{results-neo} respectively.

@@

Of all 6939 records, GPT-4 (0.61 PCCC) had the highest individual performance followed by GPT-3.5 (0.56 PCCC), InSilicoVA (0.44 PCCC), and InterVA-5 (0.44 PCCC) (Figure \ref{fig-pccc-csmf}). GPT-3.5 and GPT-4 had improvements ranging from 0.14-0.18 PCCC over InSilicoVA and InterVA-5, while GPT-4 slightly improved over GPT-3.5 by 0.05 PCCC. Population level performances were similar for all models (0.74-0.79 CSMF). Figure \ref{fig-pccc} shows the PCCC performance across three age groups (adult, child, and neonate). GPT-4 had the best individual performance for adult and neonatal records (0.64 and 0.58 PCCC), while GPT-3.5 had the best performance for child records (0.54 PCCC) with GPT-4 performing slightly worse (0.51 PCCC). InSilicoVA and InterVA-5 performed the worse for adult and child records ($\leq$0.5 PCCC), while GPT-3.5 performed the worse for neonatal records (0.42 PCCC). Across age ranges, all models followed a similar pattern in individual performance (Figure \ref{fig-pccc-age}). PCCC trended upwards for 1 month to 14 years ($\sim$0.1-0.75), and downwards for ages 15 to 69 years ($\sim$0.7-0.35). The highest and lowest performances were observed for ages 10-29 years ($\sim$) and 1-11 months respectively. Performances varied more across models for ages 0 days to 5 years, while less variation was seen between GPT-3.5 and GPT-4, as well as InSilicoVA and InterVA-5, from 5 to 69 years.

@@

Figure \ref{fig-pccc-cod-adult} shows model performance by PCCC across 17 adult CODs excluding suicide due to low sample size (n=3, \textless 1\%). GPT-4 had the highest individual performance for 10 of 17 CODs (0.35 to 0.99 PCCC), GPT-3.5 for 5 CODs (0.43-0.94 PCCC), and InSilicoVA for 2 CODs (0.71 and 0.84 PCCC). InterVA-5 had the lowest performance for 8 of 17 CODs (0-0.79 PCCC), InSilicoVA for 6 CODs (0.01-0.41 PCCC), and GPT-3.5 for 2 CODs (0.38 and 0.53 PCCC). GPT-3.5/4 models improved over InSilicoVA/InterVA-5 the most for chronic respiratory diseases (0.74-0.94 PCCC difference), and the least for Malaria (0.09-0.17 PCCC difference). All models had \textgreater 0.7 PCCC for maternal conditions (0.79-0.99 PCCC), while unspecified infections, malaria, and ill-defined CODs models with \textless 0.5 PCCC. GPT-4 had performance improvements \textgreater 0.2 PCCC compared to all other models for cancers (+0.25-0.36 PCCC), stroke (+0.27-0.45 PCCC), and diarrhoeal diseases (+0.37-0.51 PCCC), while GPT-3.5 had similar improvements for liver and alcohol related diseases (+0.27-0.52 PCCC).

@@

This section identifies the advantages of models for assigning CODs. Section \ref{discuss-adv-cod} details the application of models for particular CODs and age ranges. Section \ref{discuss-adv-hscav} details the resource efficiency of computer models for assisting in physician COD assignment. Section \ref{discuss-adv-nl} notes the strength of using natural language text in GPT models compared to structured questionnaire data for physician COD assignment.

@@

Across age ranges, all models followed a similar upward trend from 6 months to 14 years of age, and a downward trend from 15-69 years with GPT models having higher performance than InSilicoVA/InterVA-5 models, while more mixed trends were observed from 0 days to 5 months (recall Figure \ref{fig-pccc-age}). For adult age ranges, performance generally decreased as age increased, which suggested that models had difficult assigning CODs for older than younger adults with some improvements after the age of 59. For child and neonatal age ranges, the performance improved drastically as the age increased after 5 months, suggesting less difficulty in COD assignment when children and neonates are more developed. As the models did not perform particularly well ($\geq$ 0.8 PCCC) for any specific five-year age range, it is not recommended to apply specific models that target cases by age. However, the patterns of increases and decreases of performance in relation to age provide valuable insight for comparison to expected physician diagnosis patterns in well-studied medical literature and knowledge. For example, it may be expected that physicians are more uncertain in diagnosing diseases that are prevalent in neonatal patients \citep{rasmussenComplexityPhysiciansUnderstanding2019,faisonWhenUnknownUnknowable2023}, which are present in our findings from Figure \ref{fig-pccc-age}.

@@

This study evaluates the performance of GPT-3.5, GPT-4, InterVA-5, and InSilicoVA models compared to physicians for assigning CODs for 6939 VA records in Sierra Leone (2019-2022). At the population level, all models were similar (0.74-0.79 CSMF accuracy). At the individual level, GPT-4 had the best performance (0.61 PCCC), followed by GPT-3.5 (0.58 PCCC), and InSilicoVA/InterVA-5 (0.44 PCCC). Across CODs, GPT-4 had performed best for 10 of 17 adult, 4 of 8 child, and 3 of 5 neonatal CODs, with GPT-3.5 for 5 adult, 3 child, and one neonatal CODs, and InSilicoVA for 2 adult, one child, and one neonatal CODs. Model performance increased ($\sim$0.1-0.75 PCCC) as children and neonates developed (0 days to 14 years), and decreased ($\sim$0.7-0.35) as adults aged (15 to 69 years). Thus, GPT and InSilicoVA models were comparable to physicians for several CODs, but not across age ranges. As performance varied across CODs and age ranges, it is advantageous to combine several models to target CODs that each model performs well for, and to compare age-related performance patterns in relation to physicians. In addition, all models were able to scale to a large number of records and were available on-demand in comparison to physicians, enabling COD estimation and alternative diagnoses in low resource or physician scarce scenarios. As GPT models operate on natural language, they are able to adapt to more loosely defined data structures (e.g. assign in different COD coding standards, provide reasoning, and use contextual information when samples are low), making them behave more similarly to physician assignment. However, GPT models do not provide reliable CODs on repeated assignment, and were limited to past training data, with large computing infrastructure requirements, leading to reproducibility issues in COD assignments, difficulty adapting to new or changing CODs, and data privacy issues. Limitations of this study included difficulty comparing ICD-10 codes directly due to incompatible COD outputs from each model and low sample sizes, difficulty in conducting sensitivity analyses for GPT models due to costs, and omitting evaluation of multiple COD assignments due to study scope. We identified research opportunities in refining GPT models using prompt engineering and custom models for improving performance, guided household surveys to improve narrative quality, and future work in computer assisted VA, where GPT and other models will be used to assist physician COD assignment by offering multiple alternative assignments, with goals of increasing agreement on COD assignment and reducing ill-defined deaths. GPT-4, InterVA-5, and InSilicoVA has been integrated into future survey rounds of the HEAl-SL study from 2022 onwards, offering alternative COD assignments to assist physicians with second opinions. Future work in evaluating the effectiveness of computer assisted VA to reduce disagreements among physicians and ill-defined deaths will help support the advancement of more accurate and efficient VA systems across the world.

@@

This section provides additional details on the methods described in Section \ref{methods}. An overview of the methods used in this study is seen in Figure \ref{fig-methods} as a five-step process. Section \ref{app-methods-data} provides details on the preprocessed data used for modelling. Section \ref{app-methods-model} describes the data and parameter inputs and outputs for each model, while Section \ref{app-methods-eval} details the evaluation of model outputs at the individual and population level across different CODs, age groups, and age ranges.

\begin{figure}[H]
\centering
\includegraphics[width=.7\textwidth]{fig-methods.pdf}
\caption{Detailed case study methods.}\label{fig-methods}
\end{figure}

@@

The 6939 physician agreed records were collected using VA from the HEAL-SL study between 2019-2022, where records were collected using nation wide samples across Sierra Leone provinces seen in Figure \ref{fig-map}. More populous areas (e.g. southern and north east provinces with $\sim$197,000 and $\sim$135,000 population respectively) had more sampling areas versus less populous areas (e.g. north west and eastern provinces with $\sim$50,000 and $\sim$69,000 people respectively). The distribution of the case study data are shown by CGHR-10 causes of death in Table \ref{tab-data-cod}. All age groups had relatively evenly distributed female and male records (44-55\% of 6939 records each). Across CODs, there were noticeably more female records for cancers (65\%), and maternal conditions (100\%), while more male records for chronic respiratory diseases (61\%), other noncommunicable diseases (61\%), other injuries (77\%), road and transport injuries (71\%), and tuberculosis (68\%). Most records were coded by physicians as malaria for adults (20\%) and children (52\%), and stillbirth (36\%) and neonatal infections (21\%) for neonates. Suicide, congenital anomalies, nutritional deficiencies, and other had low sample sizes for each age group (\textless 1\% of total records for each age group). Table \ref{tab-data-age} shows the distribution of the study data by age ranges. Across age ranges, there were more male records for 50-59 years (60-62\%), while all other records had between 49-59\% female and male records. Most records were in the 65-69 years age range for adults (15\%), 1-5 years for children (62\%), and 0-6 days for neonates (83\%).

\begin{figure}[H]
\centering
\includegraphics[width=.75\textwidth]{fig-map.pdf}
\caption{Case study data sampling areas.}\label{fig-map}
\end{figure}

>>>>

\title[Computer Assisted Verbal Autopsy: Comparing Large Language Models to Physicians for Assigning Causes to 6939 Deaths in Sierra Leone from 2019-2022]{Computer Assisted Verbal Autopsy: Comparing Large Language Models to Physicians for Assigning Causes to 6939 Deaths in Sierra Leone from 2019-2022}

@@

\textbf{Methods:} 6939 VA records in Sierra Leone from 2019 to 2022 were used to compare four computer models, GPT-3.5, GPT-4, InterVA-5, and InSilicoVA, to physician COD assignment at population and individual levels. Narratives were used for GPT-3.5/4, while questionnaires were used for InterVA-5/InSilicoVA. COD assignments were grouped into general COD categories consisting of 19, 10, and 7 categories for adult, child, and neonatal age groups. Cause Specific Mortality Fraction (CSMF) accuracy and Partial Corrected Concordance (PCCC) were used to compare models to physicians at population and individual levels respectively. CSMF and PCCC were evaluated overall and by COD, age group, and age.

@@

\textbf{Conclusion:} All models performed similarly at the population level, while GPT-3.5/4 and InSilicoVA performed better for some CODs at the individual level. GPT models made improvements over InSilicoVA and InterVA-5. Our research lays the foundation for future work in computer assisted VA, where physicians utilize alternative COD assignments from computer models to help reduce ill-defined codes and physician disagreement.
}

@@

In 2019, 41 million people died prematurely from noncommunicable diseases every year, accounting for 74\% of all deaths globally \citep{worldhealthorganizationNonCommunicableDiseases2019}. Most of these deaths are preventable, but require adequate resource allocation, guided by evidence, to implement effective interventions and policies that target populations at risk \citep{benzigerGlobalBurdenDisease2016}. Thus, reliable counts and diagnoses of deaths enable decision makers to identify populations at risk to save lives and reduce premature deaths worldwide \citep{lawnMillionNeonatalDeaths2010,lassiCommunityBasedIntervention2015,liuExcessMortalityPersons2017,ewigCommunityacquiredPneumoniaEmergency2011}. However, many low-income countries do not have data on deaths or have registered less than half of the deaths in their country, with an even fewer 8\% of these registered deaths having a Cause of Death (COD) recorded \citep{worldhealthorganizationSCOREHealthData2021}. To fill this gap in death registrations, an alternative method known as Verbal Autopsy (VA) is used to collect data on deaths and determine their likely causes at scale \citep{desavignyIntegratingCommunitybasedVerbal2017,thomasVerbalAutopsyHealth2018,rampatigeSystematicReviewStatistics2014}, outside of traditional healthcare facilities where over half of deaths occur at home \citep{adairWhoDiesWhere2021}.

@@

Recently, Large Language Models (LLM), leveraging massive datasets and deep learning approaches, have made advances in performing a variety of Natural Language Processing (NLP) tasks using free-text, such as question answering, code generation, and even medical diagnosis \citep{changSurveyEvaluationLarge2023,lundChattingChatGPTHow2023,svyatkovskiyIntelliCodeComposeCode2020,hauptAIGeneratedMedicalAdvice2023}. In 2022, a widely-available LLM called ChatGPT was released by OpenAI with capabilities of answering natural language text inquiries using training data up to September 2021. ChatGPT-3 was based on several Generative Pre-trained Transformer (GPT) models between 2018 to 2020, namely GPT-1 to GPT-3, which had notable differences in training data sizes of 5 gigabytes to 45 terabytes from web sources that resulted in 117 million to 175 billion parameter models \citep{wuBriefOverviewChatGPT2023}. In March 2023, ChatGPT-4 was released with human-level performance on various professional and academic exams and benchmarks that outperformed ChatGPT-3 \citep{openaiGPT4TechnicalReport2023}. Given the limited usage of free-text open narratives in computer models for determining CODs, and recent advances in LLMs that leverage natural language text prompts, we conducted a study with Sierra Leone deaths from VA in 2019 to 2022 to compare four models, GPT-3.5, GPT-4, InterVA-5, and InSilicoVA, to physicians for determining CODs.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{fig-methods-brief.pdf}
\caption{Study methods.}\label{fig-methods-brief}
\end{figure}

@@

The performance of the four models were evaluated with metrics at the population and individual level by comparing their CGHR-10 COD outputs for 6939 records. Cause Specific Mortality Fraction (CSMF) accuracy was used to evaluate models on the population level (see Appendix \ref{app-methods-csmfa}), while Partial Chance Corrected Concordance (PCCC) was used to evaluate models on the individual level (see Appendix \ref{app-methods-pccc}) \citep{murrayRobustMetricsAssessing2011}. Both CSMF accuracy and PCCC metrics are between 0 and 1 with 0 indicating low performance and 1 indicating perfect performance at the population and individual level respectively. As model performance can vary across ages and specific causes \citep{benaraEvaluationMethodsAssigning2023,murrayUsingVerbalAutopsy2014,setelValidityVerbalAutopsy2006}, the CSMF accuracy and PCCC metrics were compared for each model overall, by age group (adult, child, neonatal), by CGHR-10 COD codes, and across ages. For each of the adult and child age groups, metrics were calculated for five-year ages for records with ages at death of one-year or older and five-month ages for 28 days or older. For the neonatal age group, the ages of 0-6 days and 7-27 days were used. See Appendix \ref{app-methods-eval} for more details on performance metrics and evaluation strategy for comparing each model.

@@

This section details the performance results of GPT-3.5, GPT-4, InterVA-5, and InSilicoVA models for assigning CGHR-10 CODs after applying the methods in Section \ref{methods}. GPT-4 performed the best overall at 0.61 PCCC followed by GPT-3.5 at 0.56 PCCC. GPT-4 also had the highest PCCC for most ages and CODs across the adult (12 to 69 years), child (28 days to 11 years), and neonatal (under 28 days) age groups with GPT-3.5, InterVA-5, and InSilicoVA having higher PCCC values for a few ages and CODs. Overall performance results are seen in Section \ref{results-overall}, and performance by adult, child, and neonatal records are seen in Sections \ref{results-adult}, \ref{results-child}, and \ref{results-neo} respectively.

@@

Of all 6939 records, GPT-4 (0.61 PCCC) had the highest individual performance followed by GPT-3.5 (0.56 PCCC), InSilicoVA (0.44 PCCC), and InterVA-5 (0.44 PCCC) (Figure \ref{fig-pccc-csmf}). GPT-3.5 and GPT-4 had improvements ranging from 0.14-0.18 PCCC over InSilicoVA and InterVA-5, while GPT-4 slightly improved over GPT-3.5 by 0.05 PCCC. Population level performances were similar for all models (0.74-0.79 CSMF). Figure \ref{fig-pccc} shows the PCCC performance across three age groups (adult, child, and neonate). GPT-4 had the best individual performance for adult and neonatal records (0.64 and 0.58 PCCC), while GPT-3.5 had the best performance for child records (0.54 PCCC) with GPT-4 performing slightly worse (0.51 PCCC). InSilicoVA and InterVA-5 performed the worse for adult and child records ($\leq$0.5 PCCC), while GPT-3.5 performed the worse for neonatal records (0.42 PCCC). Across ages, all models followed a similar pattern in individual performance (Figure \ref{fig-pccc-age}). PCCC trended upwards for 1 month to 14 years ($\sim$0.1-0.75), and downwards for ages 15 to 69 years ($\sim$0.7-0.35). The highest and lowest performances were observed for ages 10-29 years ($\sim$) and 1-11 months respectively. Performances varied more across models for ages 0 days to 5 years, while less variation was seen between GPT-3.5 and GPT-4, as well as InSilicoVA and InterVA-5, from 5 to 69 years.

@@

Figure \ref{fig-pccc-cod-adult} shows model performance by PCCC across 17 adult CODs excluding suicide due to low sample size (n=3, \textless 1\%). GPT-4 had the highest individual performance for 10 of 17 CODs (0.35 to 0.99 PCCC), GPT-3.5 for 5 CODs (0.43-0.94 PCCC), and InSilicoVA for 2 CODs (0.71 and 0.84 PCCC). InterVA-5 had the lowest performance for 8 of 17 CODs (0-0.79 PCCC), InSilicoVA for 6 CODs (0.01-0.41 PCCC), and GPT-3.5 for 2 CODs (0.38 and 0.53 PCCC). GPT-3.5/4 models improved over InSilicoVA/InterVA-5 the most for chronic respiratory diseases (0.74-0.94 PCCC difference), and the least for Malaria (0.09-0.17 PCCC difference). All models had \textgreater 0.7 PCCC for maternal conditions (0.79-0.99 PCCC), while \textless 0.5 PCCC for unspecified infections, malaria, and ill-defined CODs. GPT-4 had performance improvements \textgreater 0.2 PCCC compared to all other models for cancers (+0.25-0.36 PCCC), stroke (+0.27-0.45 PCCC), and diarrhoeal diseases (+0.37-0.51 PCCC), while GPT-3.5 had similar improvements for liver and alcohol related diseases (+0.27-0.52 PCCC).

@@

This section identifies the advantages of models for assigning CODs. Section \ref{discuss-adv-cod} details the application of models for particular CODs and ages. Section \ref{discuss-adv-hscav} details the resource efficiency of computer models for assisting in physician COD assignment. Section \ref{discuss-adv-nl} notes the strength of using natural language text in GPT models compared to structured questionnaire data for physician COD assignment.

@@

Across ages, all models followed a similar upward trend from 6 months to 14 years of age, and a downward trend from 15-69 years with GPT models having higher performance than InSilicoVA/InterVA-5 models, while more mixed trends were observed from 0 days to 5 months (recall Figure \ref{fig-pccc-age}). For adult ages, performance generally decreased as age increased, which suggested that models had difficult assigning CODs for older than younger adults with some improvements after the age of 59. For child and neonatal ages, the performance improved drastically as the age increased after 5 months, suggesting less difficulty in COD assignment when children and neonates are more developed. As the models did not perform particularly well ($\geq$ 0.8 PCCC) for any specific five-year age range, it is not recommended to apply specific models that target cases by age. However, the patterns of increases and decreases of performance in relation to age provide valuable insight for comparison to expected physician diagnosis patterns in well-studied medical literature and knowledge. For example, it may be expected that physicians are more uncertain in diagnosing diseases that are prevalent in neonatal patients \citep{rasmussenComplexityPhysiciansUnderstanding2019,faisonWhenUnknownUnknowable2023}, which are present in our findings from Figure \ref{fig-pccc-age}.

@@

This study evaluates the performance of GPT-3.5, GPT-4, InterVA-5, and InSilicoVA models compared to physicians for assigning CODs for 6939 VA records in Sierra Leone (2019-2022). At the population level, all models were similar (0.74-0.79 CSMF accuracy). At the individual level, GPT-4 had the best performance (0.61 PCCC), followed by GPT-3.5 (0.58 PCCC), and InSilicoVA/InterVA-5 (0.44 PCCC). Across CODs, GPT-4 had performed best for 10 of 17 adult, 4 of 8 child, and 3 of 5 neonatal CODs, with GPT-3.5 for 5 adult, 3 child, and one neonatal CODs, and InSilicoVA for 2 adult, one child, and one neonatal CODs. Model performance increased ($\sim$0.1-0.75 PCCC) as children and neonates developed (0 days to 14 years), and decreased ($\sim$0.7-0.35) as adults aged (15 to 69 years). Thus, GPT and InSilicoVA models were comparable to physicians for several CODs, but not across ages. As performance varied across CODs and ages, it is advantageous to combine several models to target CODs that each model performs well for, and to compare age-related performance patterns in relation to physicians. In addition, all models were able to scale to a large number of records and were available on-demand in comparison to physicians, enabling COD estimation and alternative diagnoses in low resource or physician scarce scenarios. As GPT models operate on natural language, they are able to adapt to more loosely defined data structures (e.g. assign in different COD coding standards, provide reasoning, and use contextual information when samples are low), making them behave more similarly to physician assignment. However, GPT models do not provide reliable CODs on repeated assignment, and were limited to past training data, with large computing infrastructure requirements, leading to reproducibility issues in COD assignments, difficulty adapting to new or changing CODs, and data privacy issues. Limitations of this study included difficulty comparing ICD-10 codes directly due to incompatible COD outputs from each model and low sample sizes, difficulty in conducting sensitivity analyses for GPT models due to costs, and omitting evaluation of multiple COD assignments due to study scope. We identified research opportunities in refining GPT models using prompt engineering and custom models for improving performance, guided household surveys to improve narrative quality, and future work in computer assisted VA, where GPT and other models will be used to assist physician COD assignment by offering multiple alternative assignments, with goals of increasing agreement on COD assignment and reducing ill-defined deaths. GPT-4, InterVA-5, and InSilicoVA has been integrated into future survey rounds of the HEAl-SL study from 2022 onwards, offering alternative COD assignments to assist physicians with second opinions. Future work in evaluating the effectiveness of computer assisted VA to reduce disagreements among physicians and ill-defined deaths will help support the advancement of more accurate and efficient VA systems across the world.

@@

This section provides additional details on the methods described in Section \ref{methods}. An overview of the methods used in this study is seen in Figure \ref{fig-methods} as a five-step process. Section \ref{app-methods-data} provides details on the preprocessed data used for modelling. Section \ref{app-methods-model} describes the data and parameter inputs and outputs for each model, while Section \ref{app-methods-eval} details the evaluation of model outputs at the individual and population level across different CODs, age groups, and ages.

\begin{figure}[H]
\centering
\includegraphics[width=.7\textwidth]{fig-methods.pdf}
\caption{Detailed study methods.}\label{fig-methods}
\end{figure}

@@

The 6939 physician agreed records were collected using VA from the HEAL-SL study between 2019-2022, where records were collected using nation wide samples across Sierra Leone provinces seen in Figure \ref{fig-map}. More populous areas (e.g. southern and north east provinces with $\sim$197,000 and $\sim$135,000 population respectively) had more sampling areas versus less populous areas (e.g. north west and eastern provinces with $\sim$50,000 and $\sim$69,000 people respectively). The distribution of the study data are shown by CGHR-10 causes of death in Table \ref{tab-data-cod}. All age groups had relatively evenly distributed female and male records (44-55\% of 6939 records each). Across CODs, there were noticeably more female records for cancers (65\%), and maternal conditions (100\%), while more male records for chronic respiratory diseases (61\%), other noncommunicable diseases (61\%), other injuries (77\%), road and transport injuries (71\%), and tuberculosis (68\%). Most records were coded by physicians as malaria for adults (20\%) and children (52\%), and stillbirth (36\%) and neonatal infections (21\%) for neonates. Suicide, congenital anomalies, nutritional deficiencies, and other had low sample sizes for each age group (\textless 1\% of total records for each age group). Table \ref{tab-data-age} shows the distribution of the study data by age. Across ages, there were more male records for 50-59 years (60-62\%), while all other records had between 49-59\% female and male records. Most records were in the 65-69 years age range for adults (15\%), 1-5 years for children (62\%), and 0-6 days for neonates (83\%).

\begin{figure}[H]
\centering
\includegraphics[width=.75\textwidth]{fig-map.pdf}
\caption{Study data sampling areas.}\label{fig-map}
\end{figure}

\begin{table}
\caption{Study data by cause of death.}\label{tab-data-cod}
\begin{tabular}{@{}lllll@{}}
    \toprule%
    Age Group  &  CGHR-10 Cause of Death (COD)  &  Female  &  Male  &  Total  \\
    \midrule
    \multirow{18}{*}{\makecell[l]{Adult, 18 CODs\\(n=3826, 55.1\%)\\Adult Female\\(n=1681, 43.9\%)\\Adult Male\\(n=2145, 56.1\%)}}  &  Acute Respiratory Infections  &  48 (45.7\%)  &  57 (54.3\%)  &  105 (2.7\%)  \\ 
        &  Cancers  &  32 (65.3\%)  &  17 (34.7\%)  &  49 (1.3\%)  \\ 
        &  Chronic Respiratory Diseases  &  29 (38.7\%)  &  46 (61.3\%)  &  75 (2\%)  \\ 
        &  Diabetes Mellitus  &  14 (51.9\%)  &  13 (48.1\%)  &  27 (0.7\%)  \\ 
        &  Diarrhoeal Diseases  &  102 (49.8\%)  &  103 (50.2\%)  &  205 (5.4\%)  \\ 
        &  Ill-Defined  &  56 (47.9\%)  &  61 (52.1\%)  &  117 (3.1\%)  \\ 
        &  Ischemic Heart Disease  &  89 (53\%)  &  79 (47\%)  &  168 (4.4\%)  \\ 
        &  Liver And Alcohol Related Diseases  &  58 (45.3\%)  &  70 (54.7\%)  &  128 (3.3\%)  \\ 
        &  Malaria  &  372 (46.6\%)  &  427 (53.4\%)  &  799 (20.9\%)  \\ 
        &  Maternal Conditions  &  130 (100\%)  &  N/A  &  130 (3.4\%)  \\ 
        &  Other Cardiovascular Diseases  &  59 (55.1\%)  &  48 (44.9\%)  &  107 (2.8\%)  \\ 
        &  Other Noncommunicable Diseases  &  160 (38.6\%)  &  254 (61.4\%)  &  414 (10.8\%)  \\ 
        &  Other Injuries  &  83 (23.2\%)  &  274 (76.8\%)  &  357 (9.3\%)  \\ 
        &  Road And Transport Injuries  &  73 (29.1\%)  &  178 (70.9\%)  &  251 (6.6\%)  \\ 
        &  Stroke  &  147 (44.4\%)  &  184 (55.6\%)  &  331 (8.7\%)  \\ 
        &  Suicide  &  N/A  &  3 (100\%)  &  3 (0.1\%)  \\ 
        &  Tuberculosis  &  54 (31.6\%)  &  117 (68.4\%)  &  171 (4.5\%)  \\ 
        &  Unspecified Infections  &  175 (45\%)  &  214 (55\%)  &  389 (10.2\%)  \\
    \midrule
    \multirow{9}{*}{\makecell[l]{Child, 9 CODs\\(n=2636, 38\%)\\Child Female\\(n=1290, 48.9\%)\\Child Male\\(n=1346, 51.1\%)}}  &  Congenital Anomalies  &  1 (100\%)  &  N/A  &  1 (0\%)  \\ 
        &  Diarrhoeal Diseases  &  79 (45.1\%)  &  96 (54.9\%)  &  175 (6.6\%)  \\ 
        &  \makecell[l]{Epilepsy, Leukaemia, And\\Other Noncommunicable Diseases}  &  61 (53.5\%)  &  53 (46.5\%)  &  114 (4.3\%)  \\ 
        &  Ill-Defined  &  34 (48.6\%)  &  36 (51.4\%)  &  70 (2.7\%)  \\ 
        &  Injuries  &  51 (37.8\%)  &  84 (62.2\%)  &  135 (5.1\%)  \\ 
        &  Malaria  &  680 (49.2\%)  &  702 (50.8\%)  &  1382 (52.4\%)  \\ 
        &  Nutritional Deficiencies  &  7 (63.6\%)  &  4 (36.4\%)  &  11 (0.4\%)  \\ 
        &  Other Infections  &  338 (50.7\%)  &  329 (49.3\%)  &  667 (25.3\%)  \\ 
        &  Pneumonia  &  39 (48.1\%)  &  42 (51.9\%)  &  81 (3.1\%)  \\
    \midrule
    \multirow{7}{*}{\makecell[l]{Neonate, 7 CODs\\(n=477, 6.9\%)\\Neonate Female\\(n=227, 47.6\%)\\Neonate Male\\(n=250, 52.4\%)}}  &  Birth Asphyxia And Birth Trauma  &  38 (36.9\%)  &  65 (63.1\%)  &  103 (21.6\%)  \\ 
        &  Congenital Anomalies  &  2 (100\%)  &  N/A  &  2 (0.4\%)  \\ 
        &  Ill-Defined  &  11 (47.8\%)  &  12 (52.2\%)  &  23 (4.8\%)  \\ 
        &  Neonatal Infections  &  49 (49.5\%)  &  50 (50.5\%)  &  99 (20.8\%)  \\ 
        &  Other  &  2 (40\%)  &  3 (60\%)  &  5 (1\%)  \\ 
        &  Prematurity And Low Birthweight  &  39 (53.4\%)  &  34 (46.6\%)  &  73 (15.3\%)  \\ 
        &  Stillbirth  &  86 (50\%)  &  86 (50\%)  &  172 (36.1\%)  \\
    \botrule
\end{tabular}
\end{table}
